---
output: html_document
editor_options: 
  chunk_output_type: console
---
Exploratory White Wine Data Analysis by Lu Tang
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='Figs/', echo=FALSE, warning=FALSE, 
                      message=FALSE)
```
# Overview & Main Question:

This project includes my thought and findings by exploring a set of wine data. 
Based on the info text, there are a few objective tests with quantitative data, 
and a column of sensory variable (taste quality) as output. By univariate, 
bivariate, and multivariate analyses with summary and reflection in the 
following report, I hope to reveal some relationship between the inputs and 
"quality". The original dataset can be found here:

> http://www3.dsi.uminho.pt/pcortez/wine/

# Quick Data Preview

```{r echo=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(GGally)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
wine <- read.csv("wineQualityWhites.csv")
```

```{r echo=FALSE}
# Get the structure and summary of the wine data frame
str(wine)
summary(wine)
```

Based on the quick analysis, the wine data frame consists of 13 variables and 
4898 observations. Among the 13 variables, the first one is the order of all 
wines tested, and the last one is output "quality". We will then focus on the 
other 11 variables impact on wine quality.

For the "quality", the actual ratings start from 1 (worst) to 9 (best), with a 
median of 6 and mean of 5.878.

Some max value are much higher than the median and 3rd quarter value 
(residual.sugar & free.sulfur.dioxide), indicating that the data set may contain
outliers. This doesn't mean those observations are dubious, they could simply be
some extreme cases. However, in order to get a good sense of the whole picture, 
the outliers are removed before analyses in each specific case.


# Univariate Plots Section

I first plot all the univariate plots, hopefully to see the distribution of each
variable.

```{r echo=FALSE, Univariate_Plots1}
create.histogram <- function(x){
  ggplot(data = wine, mapping = aes_string(x)) + 
    geom_histogram(bins = 50)
}

h.fixed.acidity <- create.histogram('fixed.acidity')
h.volatile.acidity <- create.histogram('volatile.acidity')
h.citric.acid <- create.histogram('citric.acid')
h.residual.sugar <- create.histogram('residual.sugar')
h.chlorides <- create.histogram('chlorides')
h.free.sulfur.dioxide <- create.histogram('free.sulfur.dioxide')
h.total.sulfur.dioxide <- create.histogram('total.sulfur.dioxide')
h.density <- create.histogram('density')
h.pH <- create.histogram('pH')
h.sulphates <- create.histogram('sulphates')
h.alcohol <- create.histogram('alcohol')
h.quality <- ggplot(data = wine, aes(quality)) +
  geom_histogram(bins = 7, fill = 'red')

#Show all the histogram together
grid.arrange(h.fixed.acidity, h.volatile.acidity, h.citric.acid, 
             h.residual.sugar, h.chlorides, h.free.sulfur.dioxide,
             h.total.sulfur.dioxide, h.density, h.pH,
             h.sulphates, h.alcohol, h.quality, ncol = 4)
```

Here the 12 histograms are arranged together, with the "quality" filled in 
red, since it's output. After looking through them, I find that there are a few 
parts need to be modified. Some of them has long tails due to right skewed or 
outliers, some need different binwidth to be read.

```{r echo=FALSE, Univariate_Plots1_Refined}
create.histogram2 <- function(x, variable, bw){
  ggplot(data = wine, mapping = aes_string(x)) +
    xlim(quantile(variable, 0.001),
         quantile(variable, 0.999)) +
    geom_histogram(binwidth = bw)
}

h.fixed.acidity2 <- create.histogram2('fixed.acidity', wine$fixed.acidity, 0.1)
h.volatile.acidity2 <- ggplot(wine, aes(x = volatile.acidity)) +
  xlim(quantile(wine$volatile.acidity, 0.001),
       quantile(wine$volatile.acidity, 0.999)) +
  geom_histogram() +
  scale_x_log10()
h.citric.acid2 <- create.histogram2('citric.acid', wine$citric.acid, 0.02)
h.residual.sugar2 <- ggplot(wine, aes(x = residual.sugar)) +
  xlim(quantile(wine$residual.sugar, 0.001),
       quantile(wine$residual.sugar, 0.999)) +
  geom_histogram() +
  scale_x_log10()
h.chlorides2 <- ggplot(wine, aes(x = chlorides)) +
  xlim(quantile(wine$chlorides, 0.001),
       quantile(wine$chlorides, 0.999)) +
  geom_histogram() +
  scale_x_log10()
h.free.sulfur.dioxide2 <- create.histogram2('free.sulfur.dioxide', 
                                            wine$free.sulfur.dioxide, 2)
h.total.sulfur.dioxide2 <- create.histogram2('total.sulfur.dioxide', 
                                             wine$total.sulfur.dioxide, 4)
h.density2 <- create.histogram2('density', wine$density, 0.0005)
h.pH2 <- create.histogram2('pH', wine$pH, 0.02)
h.sulphates2 <- create.histogram2('sulphates', wine$sulphates, 0.02)
h.alcohol2 <- create.histogram2('alcohol', wine$alcohol, 0.1)
h.quality2 <- ggplot(data = wine, aes(quality)) +
  xlim(quantile(wine$quality, 0.001),
       quantile(wine$quality, 0.999)) +
  geom_histogram(fill = 'red')

#Show all the histogram together
grid.arrange(h.fixed.acidity2, h.volatile.acidity2, h.citric.acid2, 
             h.residual.sugar2, h.chlorides2, h.free.sulfur.dioxide2,
             h.total.sulfur.dioxide2, h.density2, h.pH2,
             h.sulphates2, h.alcohol2, h.quality2, ncol = 4)
```

Now they look much better. Let's look into the some of the details.

```{r echo=FALSE, volatile.acidity}
grid.arrange(h.volatile.acidity, h.volatile.acidity2, ncol = 2)
```

volatile.acidity has been log transformed, with its x-axis' top and bottom 0.1%
removed. Now it's less right skewed with a bell-like shape.

```{r echo=FALSE, residual.sugar}
grid.arrange(h.residual.sugar, h.residual.sugar2, ncol = 2)
```

residual.sugar has been log transformed, with its x-axis' top and bottom 0.1%
removed. We can see a clear bimodal shape of it.

```{r echo=FALSE, chlorides}
grid.arrange(h.chlorides, h.chlorides2, ncol = 2)
```

chlorides has also been log transformed, with its x-axis' top and bottom 0.1%
removed. It also shows a sharp bell-like structure.

```{r echo=FALSE, quality}
grid.arrange(h.quality, h.quality2, ncol = 2)
```

quality is a little special. Though it seems continuous, but actually it is more
like categorical. It's range is also limited by removing top and bottom 0.1% 
outliers. With narrower bins, we can see the trend more clearly.

Notice here I set the upper and lower limit really strict: only top and bottom 
0.1% of the data are treated as outliers. Let's see what would happen if we set 
the limit to 1% for "quality".

```{r echo=FALSE, WQ}
table(wine$quality == 3)
```

Notice there are only 20 observations are equal to 3, which is smaller than 1% 
of 4878. The RStudio will then remove all the values equal to 4 since it doesn't
know which of the "4s" to keep, so that we can only see values from 5 in 
"quality" plot. Similar phenomena happens to the "density" plots. This could 
probably mean that the white wine data set doesn't have much outliers.

# Univariate Analysis

### What is the structure of your dataset?

There are 4898 white wines in the dataset with 11 quatitative featuresand 1 
output variable (sensory "quality" between 0 to 10, but the actual values range 
from 3 to 9). The mean quality is 5.878 and median is 6.

### What is/are the main feature(s) of interest in your dataset?

The main features of interest are how these input variables related to the wine 
"quality", or how some of them together correlated with the output (i.e. maybe 
citric acid + residual sugar + total sulful dioxide together strongly impact the
wine quality).

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I think it is better to look into these variables by several groups: acidity, 
sulfur and the other, then try to see if there exist a correlation between 
groups.

### Did you create any new variables from existing variables in the dataset?

Not for univariate analysis yet.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

After plotting all the univariate features into histogram, I found that many of 
them are skewed/long tailed. So I removed the outliers of top and bottom 0.1% of 
each variable, and changed some of the x-axis into log form. I also adjusted the
bin width for each of the histograms.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots, dpi = 600}
head(wine)
ggpairs(wine, columns = 2:13,
        lower = list(continuous = wrap('points', alpha = 0.5, size = 0.5)),
        upper = list(continuous = wrap('cor', size = 3)),
        axisLabels = 'show') +
  theme(axis.text = element_text(size = 6))
```

I double checked the wine dataset by listing its first 6 observations for each 
column. The first one is an order factor, hence I plotted the other 12 
variables' relationship into a matrix plot. We can get a quick and brief idea 
about the data frame. For example, it is easy to tell that residual.sugar has a 
positive correlation with density. Then I will pick a few of those variables for
further exploration.

In the matrix, I used all the 4898 observations for each of the variables, so 
the correlation are more precise than taking random samples. There are 3 parts 
in the matrix plot: the lower left part shows all the bivariate point plots, 
while their corresponding correlation are shown on upper right part, and the 
diagonal plots are the density function of each of the variables.

We can easily tell some strong bivariate correlation either from high absolute 
value of "Corr", or their dense points distribution. For our main interest about
quality, the volatile.acid has stronger correlation (-0.195) than fixed.acid 
(-0.114), and the citric.acid almost not related to quality (-0.00921). On the 
other hand, chlorides (-0.21), density (-0.307) and alcohol (0.436) play more
important roles on quality. 

At the same time, correlation between some input variables are also notable and 
quite interesting. For example, it is easy to imply a strong correlation between free.sulfur.dioxide and total.sulfur.dioxide (0.616), since the former is a part
of the latter. The most significant correlation is between residual.sugar to 
density (0.839), which is almost linear. As mentioned before, density has a 
strong correlation with quality, however this is not true for residual.sugar 
(-0.0976), which is quite interesting to dig into. I also noticed a strong 
correlation between density and alcohol (-0.78), and both of them are strongly 
correlated with quality. 

I may not explain the above phenomena in this report without the help of 
statistical inference. Nonetheless it won't block me to take a close look at 
these bivariate relationship.

Firstly, I chose a few input variables with output "quality". I picked 
residual.sugar as not significant related, and chlorides, density and alcohol as 
more related to "quality". 

```{r}
ggplot(data = wine, aes(x = quality, y = residual.sugar)) + 
  ylim(c(0, 15)) +
  geom_jitter(alpha = 0.1) +
  geom_line(stat = "summary", fun.y = mean, color = "red")
```

```{r}
ggplot(data = wine, aes(x = quality, y = residual.sugar, group = quality)) + 
  ylim(c(0, 15)) +
  geom_boxplot()
```

The residual.sugar median for each quality goes irregularly. At the same 
time, each 1st to 3rd quantile of residual.sugar covers a long range and 
overlapping with the others. So there is no trend to tell whether high or low 
residual.sugar leads to a good quality. 

```{r}
ggplot(data = wine, aes(x = quality, y = chlorides)) + 
  ylim(c(0.01, 0.06)) +
  geom_jitter(alpha = 0.1) +
  geom_line(stat = "summary", fun.y = mean, color = "red")
```

```{r}
ggplot(data = wine, aes(x = quality, y = chlorides, group = quality)) +
  ylim(c(0.01, 0.06)) +
  geom_boxplot()
```

```{r}
ggplot(data = wine, aes(x = quality, y = density)) + 
  ylim(c(0.988, 1)) +
  geom_jitter(alpha = 0.1) +
  geom_line(stat = "summary", fun.y = mean, color = "red")
```

```{r}
ggplot(data = wine, aes(x = quality, y = density, group = quality)) +
  ylim(c(0.988, 1)) +
  geom_boxplot()
```

```{r}
ggplot(data = wine, aes(x = quality, y = alcohol)) + 
  ylim(c(9, 13)) +
  geom_jitter(alpha = 0.1) +
  geom_line(stat = "summary", fun.y = mean, color = "red")
```

```{r}
ggplot(data = wine, aes(x = quality, y = alcohol, group = quality)) +
  ylim(c(9, 13)) +
  geom_boxplot()
```

On the other hand, the other 3 set of plots all showed some trends through the 
median, though they are not monotonic. They all have inflection points at 
quality of 5. For chlorides and density, the main trend after inflection points 
are decreasing with the increasement of quality, proving their negative 
correlation. However, it is not enough to guarantee good quality by any single 
variable of them. In each boxplot, at quality of 4 and 6 all have a lot of 
overlaps. This leads to an analysis of multivariate relatioship in latter part 
of this report.

Meanwhile, let's take a look at the correlation between input variables.

```{r echo=FALSE, Bivariate_Plots3}
bp.residual.sugar.density <-
  ggplot(data = wine, aes(x = residual.sugar, y = density)) +
  geom_jitter(alpha = 0.2, color = "red") +
  xlim(0, quantile(wine$residual.sugar, 0.95)) +
  ylim(0.987, quantile(wine$density, 0.95)) +
  geom_smooth()

bp.residual.sugar.density
```

As expected, the density is strongly correlated with residual.sugar. But it 
looks like three distinct part with different slopes, especially at 
residual.sugar = 3. Remember in the univariate analysis part, only 
residual.sugar shows a bimodal distribution. Hence I'd like to check the 
pearson correlation at the two different part.

```{r echo=FALSE, Cor.test}
wine.rs1 <- subset(wine, residual.sugar <= 3)
wine.rs2 <- subset(wine, residual.sugar > 3)
print("correlation of density and residual.sugar <= 3")
cor.test(wine.rs1$density, wine.rs1$residual.sugar)
print("correlation of density and residual.sugar > 3")
cor.test(wine.rs2$density, wine.rs2$residual.sugar)
```

It is quite interesting to see that they behaved so differently: the former part
is almost non-correlated, but the latter is strongly correlated. It tells that:
the density is less determined by residual.sugar at low residual.sugar level, 
but they are highly correlated at high residual.sugar level.

```{r echo=FALSE, Bivariate_Plots4}
bp.alcohol.density <-
  ggplot(data = wine, aes(x = alcohol, y = density)) +
  geom_jitter(alpha = 0.2, color = "red") +
  xlim(7.5, 14.5) +
  ylim(0.987, quantile(wine$density, 0.95)) +
  geom_smooth()

bp.alcohol.density
```

Also, density is strongly negatively correlated with alcohol.

# Bivariate Analysis

In the bivariate analysis part, I first checked the matrix plot to find out some
potential interesting relationship. Then I studied 4 input variables's 
correlation with output "quality" with boxplots. They are residual.sugar, 
chlorides, density, andclcohol. I also checked the relationships between two 
sets of input features: residual.sugar to density and alcohol to density.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the data set?

Since our main interst is to find out if there is a feature that lead to a high 
"quality", the bivariate section gives us a good opportunity to check them. As 
shown in the boxplots, residual.sugar is a very bad indicator to quality, while 
chlorides, density, and alcohol show relatively strong correlation with quality.
However, none of them alone can be used to predict quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Density has strong positive correlation with residual.sugar (0.839) and negative
correlation with alcohol (-0.78). From the points and smooth line plots, we can 
see almost linear relationship. This is quite interesting since both density and
alcohol also correlated with quality. In my opinion, alcohol is lighter than 
water, hence more alcohol lead to lower density. If we use alcohol as a feature 
to predict quality, we should not use density together.

### What was the strongest relationship you found?

The strongest relationship in my bivariate analysis is between residual.sugar 
with density (0.839). This is reasonable physically.


# Multivariate Plots Section

In the previous analysis, I found that chlorides and alcohol have relatively 
significant correlation with quality. So I put these 3 together, marking quality
with different colors.

```{r echo=FALSE, Multivariate_Plots1}
ggplot(data = wine, aes(x = alcohol, y = chlorides)) +
  geom_point(aes(color = factor(quality))) +
  scale_color_brewer(type = "qual", palette = "Blues")
```

We can see a clear trend that higher alcohol indicating higher quality. However,
we cannot tell how is chlorides correlated with quality. In other words, as a 
multivariate plot, it doesn't give us more information than a bivariate plot of 
alcohol to quality.

I'm then interested to see the relationship among density, residual.sugar and 
alcohol. As discussed above, these three features are strongly related.

```{r echo=FALSE, Multivariate_Plots2}
wine.rs <- subset(wine, residual.sugar <= quantile(wine$residual.sugar, 0.95))
ggplot(data = wine.rs, aes(x = alcohol, y = residual.sugar)) +
  geom_point(aes(color = density)) +
  scale_color_gradientn(colours = rainbow(5))
```

Great! A very well separated rainbow pattern appears. We can easily see that: 
higher residual.sugar and lower alcohol lead to higher density. This proves my
previous analysis. 

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

My main interest is to find a few features that can work together to predict 
quality, thus I picked the most significant correlated features with quality. 
However, they don't work together as discussed in the first mulitvariate plots.
At the same time, I found a pretty good correlation among density, 
residual.sugar and alcohol.

### Were there any interesting or surprising interactions between features?

The residual.sugar and alcohol can be good indicator for density. Because of 
this, density should not be used together with alcohol to predict quality.


------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
bp.q.alcohol2 <-
  ggplot(data = wine, aes(x = quality, y = alcohol, group = quality)) +
  ylim(c(9, 13)) +
  ylab("alcohol (% by volume)") +
  ggtitle("Boxplot Alcohol vs Quality") +
  geom_boxplot()

bp.q.alcohol2
```

### Description One

The boxplot shows relatively strong positive correlation between alcohol and 
quality. Although the relationship is not linear, we can still see a clear trend
especially with quality > 5. However, the overlapping made it's not reliable to
predict quality with alcohol as single input variable.

Along with the other boxplots, these very basic correlation can help me to find 
out some interesting features that are relatively strongly related to quality.

### Plot Two
```{r echo=FALSE, Plot_Two}
bp.alcohol.density2 <-
  ggplot(data = wine, aes(x = alcohol, y = density)) +
  geom_jitter(alpha = 0.2, color = "red") +
  xlim(7.5, 14.5) +
  ylim(0.987, quantile(wine$density, 0.95)) +
  xlab("alcohol (% by volume)") +
  ylab("density (g / cm^3)") +
  ggtitle("Relation between Density and Alcohol") +
  geom_smooth()

bp.alcohol.density2
```

### Description Two

Density is strongly negatively correlated with alcohol. In this case, if we use
both density and alcohol to predict quality, yes, we will get a good 
correlation. However, it is meaningless. 

### Plot Three
```{r echo=FALSE, Plot_Three}
wine.rs2 <- subset(wine, residual.sugar <= quantile(wine$residual.sugar, 0.95))
ggplot(data = wine.rs2, aes(x = alcohol, y = residual.sugar)) +
  geom_point(aes(color = density)) +
  scale_color_gradientn(colours = rainbow(5)) +
  xlab("alcohol (% by volume)") +
  ylab("residual sugar (g / dm^3)") +
  ggtitle("Density Distribution by Residual Sugar and Alcohol") +
  labs(color = "density (g / cm^3)")
```

### Description Three

Instead of predicting of quality by multi input features, density distribution 
can be perfectly predicted by residual sugar and alcohol. A nice rainbow like 
pattern indicates that higher residual sugar and lower alcohol leads to higher
density, and vice versa.


------

# Reflection

This is my first exploratory data analysis project. By now, I learnt to use 
extensive libraries in R. I got familiar with checking data information through 
dim, str, ploting with multiple types, adjusting dataset and removing outliers, 
and improving charts' readability through changing transparency, jitter, 
smoothing, labeling and so on. Most importantly, all the above are done with R 
markdown. I'm now confident with the tool to create professional-looking 
reports.

Another great thing I learnt is doing analysis step by step. We need to go 
through univariate analysis to bivariate, then to multivariate analysis. In each
of them, it's good to have a big picture rather than randomly picking any 
feature. I plotted all the 12 univariate plots, and all bivariate plots in a 
matrix plot. They help me a lot to go through this project.

Originally, I would like to find some features that can working together to give
a good prediction of output "quality", however not able to do it without the 
help of statistical inference or model training. I can't wait to keep learning.